\large How resource abundance and stochasticity affect animals' space-use requirements

### 2023-11-23 writing/appendix-C-empirical-modeling.Rmd
prepared manuscript and appendices for submission

---
title: \large How resource abundance and stochasticity affect animals' space-use requirements
subtitle: "Appendix 3: Empirical modeling"
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: [biol, braes]
  - name: Chris H. Fleming
    institute: [ucf, smith]
  - name: E. Patrícia Medici
    institute: [tapir, iucn, escas]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [biol, braes, cmps]
institute:
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
  - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
  - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
  - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'hr-resource-stoch.bib'
csl: 'the-american-naturalist.csl'
fontsize: 12pt
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize} # smaller caption text
    -  \usepackage{hanging} # for hanging indents in references
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
---

@@

This appendix illustrates the steps necessary to reproduce the tapir movement analysis and the related figure in the manuscript. The tapir data used here is from the work of @medici_movement_2022 and can be found at the GitHub repository located at [https://github.com/StefanoMezzini/tapirs](https://github.com/StefanoMezzini/tapirs). For ease of reference, we also include the figure below (fig. \@ref(fig:tapir-mw)). Estimating the effects of resource abundance and unpredictability on the tapir's space-use requirements requires us to first estimate the changes in the tapir's space-use requirements (section \@ref(movement), fig. \@ref(fig:tapir-mw)c) and the changes in resource abundance and variance experienced by the tapir (section \@ref(ndvi), figs. \@ref(fig:tapir-mw)a-b) before we can estimate the relationship between resource dynamics and spatial use (section \@ref(final-fig), figs. \@ref(fig:tapir-mw)d-f).

```{r tapir-mw, fig.cap="Seven-day home-range size of a lowland tapir (\\emph{Tapirus terrestris}) in response to changes in mean and variance in resource abundance. (a.) Mean resource abundance, $\\mu(t)$, esimated as the mean NDVI at the locations visited by the tapir. (b.) Varince in resource abundance, $\\sigma^2(t)$, esimated as the average variance in NDVI at the locations visited by the tapir. (c.) Estimated seven-day home range based on the 95\\% utilization quantiles. (d., e.) estimated marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ on home-range size. The model accounted for the marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ and their interaction on mean space-use requirements and the variance around them. (f.) The effect of $\\mu(t)$ does not follow the data closely because $\\e(R)$ and $\\var(R)$ are highly correlated. Consequently, while estimating the effects of $\\e(R)$ and $\\var(R)$ via separate models would allow result in a closer fit, the estimated effects would be inappropriate because they do not disentangle the effects of $\\e(R)$ and $\\var(R)$. See Appendix 3 for additional information. The tapir movement data corresponds to the individual named ``Anna\" from the Cerrado sample of Medici $et~al.$ (2022).", out.width='100%', eval=TRUE, echo=FALSE}
knitr::include_graphics('../figures/tapir-example-with-data.png', rel_path = FALSE)
```

@@

```{r, eval=TRUE}
getwd() # change working directory or file paths if needed

# attach all necessary packages
library('terra')    # to import and save rasters
library('dplyr')     # for data wrangling
library('purrr')     # for functional programming
library('tidyr')     # for data wrangling
library('ggplot2')   # for fancy plots
library('cowplot')   # for fancy multi-panel plots
library('ctmm')      # for movement modeling
library('mgcv')      # for empirical Bayesian GAMs
library('scam')      # for shape-constrained splines
library('lubridate') # for smoother date wrangling
library('sf')        # for spatial features
library('MODIStsp')  # for downloading NDVI rasters
theme_set(theme_bw()) # change default theme

@@

The `window_hr()` function estimates the tapir's home range using a sliding window approach with a window size of 7 days (using `ctmm`'s `%#%` operator for unit conversions) that starts with the set of days from `r format(as.POSIXct(anna_mw$t_start[1]), format = '%Y-%m-%d')` to `r format(as.POSIXct(anna_mw$t_end[1]), format = '%Y-%m-%d')` (extremes included), then shifts by one day (`dt = 1 %#% 'day'`), repeats the analysis for the next seven-day set (`r format(as.POSIXct(anna_mw$t_start[2]), format = '%Y-%m-%d')` to `r format(as.POSIXct(anna_mw$t_end[2]), format = '%Y-%m-%d')`), and continues doing so until it reaches the last set of days, `r format(as.POSIXct(anna_mw$t_start[nrow(anna_mw)]), format = '%Y-%m-%d')` to `r format(as.POSIXct(anna_mw$t_end[nrow(anna_mw)]), format = '%Y-%m-%d')`. For each set of days, it fits a positional variogram, a continuous-time movement model [@fleming_ctmm_2021], and a utilization distribution via autocorrelated kernel density estimation [@silva_autocorrelationinformed_2022; @noonan_comprehensive_2019]. Finally, it saves an exploratory figure (fig. \@ref(fig:mw)) to the `figures` folder and the tibble of times, telemetries, movement models, utilization distributions, and home range estimates (with 95% confidence intervals) to the `models` folder.

```{r mw, eval=TRUE, echo=FALSE, fig.cap="Exploratory figure created by the $\\tt{window\\_hr()}$ function. Panel $\\mathbf{a.}$ shows the tapir's GPS locations, while panel $\\mathbf{b.}$ shows the seven-day home range estimates (95\\% and 50\\% utilization quantiles) with 95\\% confidence intervals.", out.width='\\textwidth'}
knitr::include_graphics('../figures/tapirs/CE_31_ANNA-window-7-days-dt-1-days.png')
```

@@

```{r, echo=FALSE, eval=TRUE}
# import NDVI data
anna_ndvi <-
  readRDS('data/ndvi-rasters/tapir-anna/tapir-anna-data.rds') %>%
  mutate(dec_date = decimal_date(date))
anna_ndvi
```

@@

We fit the beta GAMLS via the `mgcv` package (note `family = betals()` in the code chunk below). The `betals` family accepts a list of two predictors: one for the mean parameter, $\mu$, and one for the scale parameter, $\phi$, and it uses logit link functions for both parameters (see fig. \@ref(fig:gamls-terms)). The variance of the distribution is a function of both parameters: $\sigma^2 = \mu (1 - \mu) \phi$.

@@

Note that when fitting location-scale GAMs, one should pay particular attention to the number of knots used for each smooth term. While using a penalized maximum likelihood method such as REML (`method = 'REML'`) helps avoid over-fitting the model, finding the right balance between each of the `k` values is crucial. Excessively smooth terms for the mean can inflate the scale term (fig. \@ref(fig:undersmooth)), while excessively wiggly terms for the mean can cause the scale to be under-estimated (fig. \@ref(fig:oversmooth)). Ultimately, each of the `k` values should be decided in such a way as to mimic the animal's responsiveness, adaptability, motility, and memory (or ability to predict cycles or events). If one is unsure where to start from, keeping the `k` for the scale terms below half the `k` for the mean terms is a good starting point.

@@

```{r gamls-terms, echo=FALSE, eval=TRUE, fig.height=8, fig.cap="Estimated spatiotemporal trends in mean and scale parameters using the model detailed in the code chunk above. Estimates are provided on the logit link scale. The estimated degrees of freedom for each term can be seen in parentheses in the title of the spatial terms and the y-axis labels of the temporal terms. Dashed lines inticate the 95\\% credible intervals for the temporal terms.", fig.pos='!h'}
m_ndvi <- readRDS('models/tapirs/CE_31_ANNA-mgcv-ndvi-betals.rds')
plot(m_ndvi, pages = 1, scheme = 3, n = 250, scale = 0)
```

@@

```{r}
tapir <-
  readRDS('models/tapirs/CE_31_ANNA-window-7-days-dt-1-days.rds') %>%
  mutate(est = map(dataset,
                   \(.d) filter(anna_tel, timestamp %in% .d$timestamp)),
         mu = map_dbl(est, \(.d) mean(.d$mu)),
         sigma2 = map_dbl(est, \(.d) mean(.d$sigma2))) %>%
  select(date, mu, sigma2, hr_lwr_95, hr_est_95, hr_upr_95) %>%
  mutate(weight = 1 / (hr_upr_95 - hr_lwr_95), # small CI => large weight
         weight = weight / mean(weight)) # standardize weights
sum(tapir$weight) == nrow(tapir) # ensure weights are right
```

```{r, eval=TRUE, echo=FALSE}
tapir <- readRDS('data/anna-hr-ndvi-data.rds')
sum(tapir$weight) == nrow(tapir) # ensure weights are right
```

We now have all we need to create the left side of figure \@ref(fig:tapir-mw).

@@

l_grobs <- lapply(
  list(ggplot(tapir, aes(date, mu)) + # mean
         geom_line(color = pal[1]) +
         labs(x = NULL, y = e_r),
       ggplot(tapir, aes(date, sigma2)) + # variance
         geom_line(color = pal[2]) +
         labs(x = NULL, y = v_r),
       ggplot(tapir, aes(date, hr_est_95)) + # 95% home range
         geom_line(color = pal[3]) +
         labs(x = NULL, y = hr_lab)),
  as_grob) # convert to grid graphical objects (grobs)

@@

# Draw aligned plots
plot_grid(plotlist = l_grobs, nrow = 1, labels = c('a.', 'b.', 'c.'))
```

@@

To create the right side of the figure, we will need to estimate the effects of $\e(R)$ and $\var(R)$ on the tapir's space use. To do this, we fit a GAM to the the tapir's 7-day home range estimates using the mean and variance in NDVI as predictors. However, since the two predictors are highly correlated, using a regular GAM causes the effects to be non-monotonic (after strongly restricting `k` to avoid over-fitting; see fig. \@ref(fig:hr-gam)).

@@

plot_grid(plotlist = r_grobs, ncol = 3, labels = c('d.', 'e.', 'f.'))
```

>>>>

---
title: \large How resource abundance and stochasticity affect animals' space-use requirements
subtitle: "Appendix C: Empirical modeling"
# author:
#   - name: Stefano Mezzini
#     email: stefano.mezzini@ubc.ca
#     institute: [biol, braes]
#   - name: Chris H. Fleming
#     institute: [ucf, smith]
#   - name: E. Patrícia Medici
#     institute: [tapir, iucn, escas]
#   - name: Michael J. Noonan
#     email: michael.noonan@ubc.ca
#     institute: [biol, braes, cmps]
# institute:
#   - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
#   - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
#   - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
#   - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
#   - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
#   - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
#   - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
#   - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'hr-resource-stoch.bib'
csl: 'the-american-naturalist.csl'
fontsize: 12pt
header-includes:
    - \renewcommand{\figurename}{Figure C\!\!} # for "Figure Sx.
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize} # smaller caption text
    -  \usepackage{hanging} # for hanging indents in references
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
---

@@

This appendix illustrates the steps necessary to reproduce the tapir movement analysis and the related figure in the manuscript (fig. 5). The tapir data used here is from the work of @medici_movement_2022 and can be found at the GitHub repository located at [https://github.com/StefanoMezzini/tapirs](https://github.com/StefanoMezzini/tapirs). Estimating the effects of resource abundance and unpredictability on the tapir's space-use requirements requires us to first estimate the changes in the tapir's space-use requirements (section \@ref(movement)) and the changes in resource abundance and variance experienced by the tapir (section \@ref(ndvi)) before we can estimate the relationship between resource dynamics and spatial use (section \@ref(final-fig)).

@@

```{r, eval=FALSE}
getwd() # change working directory or file paths if needed
```

```{r, eval=TRUE, echo=FALSE}
print('H:/GitHub/hr-resource-stoch/writing') # for anonymity in peer review
```

```{r, eval=TRUE}
# attach all necessary packages
library('terra')    # to import and save rasters
library('dplyr')     # for data wrangling
library('purrr')     # for functional programming
library('tidyr')     # for data wrangling
library('ggplot2')   # for fancy plots
library('cowplot')   # for fancy multi-panel plots
library('ctmm')      # for movement modeling
library('mgcv')      # for empirical Bayesian GAMs
library('scam')      # for shape-constrained splines
library('lubridate') # for smoother date wrangling
library('sf')        # for spatial features
library('MODIStsp')  # for downloading NDVI rasters
theme_set(theme_bw()) # change default theme

@@

The `window_hr()` function estimates the tapir's home range using a sliding window approach with a window size of 7 days (using `ctmm`'s `%#%` operator for unit conversions) that starts with the set of days from `r format(as.POSIXct(anna_mw$t_start[1]), format = '%Y-%m-%d')` to `r format(as.POSIXct(anna_mw$t_end[1]), format = '%Y-%m-%d')` (extremes included), then shifts by one day (`dt = 1 %#% 'day'`), repeats the analysis for the next seven-day set (`r format(as.POSIXct(anna_mw$t_start[2]), format = '%Y-%m-%d')` to `r format(as.POSIXct(anna_mw$t_end[2]), format = '%Y-%m-%d')`), and continues doing so until it reaches the last set of days, `r format(as.POSIXct(anna_mw$t_start[nrow(anna_mw)]), format = '%Y-%m-%d')` to `r format(as.POSIXct(anna_mw$t_end[nrow(anna_mw)]), format = '%Y-%m-%d')`. For each set of days, it fits a positional variogram, a continuous-time movement model [@fleming_ctmm_2021], and a utilization distribution via autocorrelated kernel density estimation [@silva_autocorrelationinformed_2022; @noonan_comprehensive_2019]. Finally, it saves an exploratory figure (fig. C\@ref(fig:mw)) to the `figures` folder and the tibble of times, telemetries, movement models, utilization distributions, and home range estimates (with 95% confidence intervals) to the `models` folder.

```{r mw, eval=TRUE, echo=FALSE, fig.cap="Exploratory figure created by the $\\tt{window\\_hr()}$ function. Panel $\\mathbf{A}$ shows the tapir's GPS locations, while panel $\\mathbf{B}$ shows the seven-day home range estimates (95\\% utilization quantile) with 95\\% confidence intervals.", out.width='\\textwidth'}
knitr::include_graphics('../figures/tapirs/CE_31_ANNA-window-7-days-dt-1-days.png')
```

@@

```{r, echo=FALSE, eval=TRUE}
# import NDVI data
anna_ndvi <-
  readRDS('../data/ndvi-rasters/tapir-anna/tapir-anna-data.rds') %>%
  mutate(dec_date = decimal_date(date))
anna_ndvi
```

@@

We fit the beta GAMLS via the `mgcv` package (note `family = betals()` in the code chunk below). The `betals` family accepts a list of two predictors: one for the mean parameter, $\mu$, and one for the scale parameter, $\phi$, and it uses logit link functions for both parameters (see fig. C\@ref(fig:gamls-terms)). The variance of the distribution is a function of both parameters: $\sigma^2 = \mu (1 - \mu) \phi$.

@@

Note that when fitting location-scale GAMs, one should pay particular attention to the number of knots used for each smooth term. While using a penalized maximum likelihood method such as REML (`method = 'REML'`) helps avoid over-fitting the model, finding the right balance between each of the `k` values is crucial. Excessively smooth terms for the mean can inflate the scale term (fig. C\@ref(fig:undersmooth)), while excessively wiggly terms for the mean can cause the scale to be under-estimated (fig. C\@ref(fig:oversmooth)). Ultimately, each of the `k` values should be decided in such a way as to mimic the animal's responsiveness, adaptability, motility, and memory (or ability to predict cycles or events). If one is unsure where to start from, keeping the `k` for the scale terms below half the `k` for the mean terms is a good starting point.

@@

```{r gamls-terms, echo=FALSE, eval=TRUE, fig.height=8, fig.cap="Estimated spatiotemporal trends in mean and scale parameters using the model detailed in the code chunk above. Estimates are provided on the logit link scale. The estimated degrees of freedom for each term can be seen in parentheses in the title of the spatial terms and the y-axis labels of the temporal terms. Dashed lines inticate the 95\\% credible intervals for the temporal terms.", fig.pos='!h'}
m_ndvi <- readRDS('../models/tapirs/CE_31_ANNA-mgcv-ndvi-betals.rds')
plot(m_ndvi, pages = 1, scheme = 3, n = 250, scale = 0)
```

@@

```{r}
tapir <-
  readRDS('../models/tapirs/CE_31_ANNA-window-7-days-dt-1-days.rds') %>%
  mutate(est = map(dataset,
                   \(.d) filter(anna_tel, timestamp %in% .d$timestamp)),
         mu = map_dbl(est, \(.d) mean(.d$mu)),
         sigma2 = map_dbl(est, \(.d) mean(.d$sigma2))) %>%
  select(date, mu, sigma2, hr_lwr_95, hr_est_95, hr_upr_95) %>%
  mutate(weight = 1 / (hr_upr_95 - hr_lwr_95), # small CI => large weight
         weight = weight / mean(weight)) # standardize weights
sum(tapir$weight) == nrow(tapir) # ensure weights are right
```

```{r, eval=TRUE, echo=FALSE}
tapir <- readRDS('../data/anna-hr-ndvi-data.rds')
sum(tapir$weight) == nrow(tapir) # ensure weights are right
```

We now have all we need to create the left side of figure 5 from the main manuscript.

@@

l_grobs <- lapply(
  list(ggplot(tapir, aes(date, mu)) + # mean
         geom_line(color = pal[1], size = 2) +
         labs(x = NULL, y = e_r),
       ggplot(tapir, aes(date, sigma2)) + # variance
         geom_line(color = pal[2], size = 2) +
         labs(x = NULL, y = v_r),
       ggplot(tapir, aes(date, hr_est_95)) + # 95% home range
         geom_line(color = pal[3], size = 2) +
         labs(x = NULL, y = hr_lab)),
  as_grob) # convert to grid graphical objects (grobs)

@@

# Draw aligned plots
plot_grid(plotlist = l_grobs, nrow = 1, labels = 'AUTO')
```

@@

To create the right side of the figure, we will need to estimate the effects of $\e(R)$ and $\var(R)$ on the tapir's space use. To do this, we fit a GAM to the the tapir's 7-day home range estimates using the mean and variance in NDVI as predictors. However, since the two predictors are highly correlated, using a regular GAM causes the effects to be non-monotonic (after strongly restricting `k` to avoid over-fitting; see fig. C\@ref(fig:hr-gam)).

@@

plot_grid(plotlist = r_grobs, ncol = 3, labels = c('D', 'E', 'F'))
```
